Optimizing and Incrementalizing Collection Queries by AST Transformations

### 2017-06-04 aosd13/paperBody.tex

\subsection{LMS}
Our deep embedding is inspired by some of the Scala techniques presented by
Lightweight Modular Staging (LMS)~\citep{rompf2010lightweight} for using
implicits and for adding infix operators to a type. Like
\citet{rompf2010lightweight}, we also generate and compile Scala code on-the-fly
reusing the Scala compiler. A plausible alternative backend for \LoS\ would have
been to use LMS and Delite~\citep{Rompf11BBlocks}, a framework for building
highly efficient DSLs in Scala.

@@

OptiQL achieves impressive speedups by fusing collection operations and
transforming (or \emph{lowering}) high-level bulk operators to highly optimized
imperative programs using \texttt{while} loops. This gains significant
advantages because it can avoid boxing, intermediate allocations, and because
inlining heuristics in the Hotspot JIT compiler have never been tuned to bulk
operators in functional programs.\footnote{See discussion by Cliff Click at \url{http://www.azulsystems.com/blog/cliff/2011-04-04-fixing-the-inlining-problem}.}\pg{Maybe elaborate and cite Cliff Click's blog post, turn it into a citation?}
We did not investigate such lowerings. It is unclear how
well these optimizations would extend to other collections which intrinsically
carry further overhead.
\pg{Resume and connect with \cref{sec:optim-vs-inc}.}
\pg{Cite Stream fusion to completeness~\citep{Kiselyov2017stream}!}

>>>>

\subsection{LMS}
\label{sec:rwdsl-lms}
Our deep embedding is inspired by some of the Scala techniques presented by
Lightweight Modular Staging (LMS)~\citep{rompf2010lightweight} for using
implicits and for adding infix operators to a type. Like
\citet{rompf2010lightweight}, we also generate and compile Scala code on-the-fly
reusing the Scala compiler. A plausible alternative backend for \LoS\ would have
been to use LMS and Delite~\citep{Rompf11BBlocks}, a framework for building
highly efficient DSLs in Scala.

@@

OptiQL achieves impressive speedups by fusing collection operations and
transforming (or \emph{lowering}) high-level bulk operators to highly optimized
imperative programs using \texttt{while} loops. This gains significant
advantages because it can avoid boxing, intermediate allocations, and because
inlining heuristics in the Hotspot JIT compiler have never been tuned to bulk
operators in functional programs.\footnote{See discussion by Cliff Click at \url{http://www.azulsystems.com/blog/cliff/2011-04-04-fixing-the-inlining-problem}.}\pg{Maybe elaborate and cite Cliff Click's blog post, turn it into a citation?}
We did not investigate such lowerings. It is unclear how
well these optimizations would extend to other collections which intrinsically
carry further overhead. Moreover, it is unclear how to execute incrementally the
result of these lowerings, as we discuss in \cref{sec:optim-vs-inc}.
\pg{Cite Stream fusion to completeness~\citep{Kiselyov2017stream}!}

### 2017-06-04 aosd13/pepm13/caseStudyDiscussion.tex

\section{Optimization limitations}
\label{sec:optim-vs-inc}
\pg{Resume}
While indexing achieved significant speedups, when indexing does not apply
speedups are more limited; in comparison, later projects working on collection
query optimization gave better speedups.

@@

Another problem is that most fusion techniques are designed for sequential processing, hence conflict with incrementalization.
Most fusion techniques generally assume bulk operations scan collections in
linear order. For instance, shortcut fusion rewrites operators in terms of
\texttt{foldr}. It is not so clear how to extend them to parallel and
incremental programs. \pg{Cite Guy Steele talk against cons, and Ingo Maier's
  incremental library.}

>>>>

\section{Optimization limitations}
\label{sec:optim-vs-inc}
In our experiments indexing achieved significant speedups, but when indexing
does not apply speedups are more limited; in comparison, later projects working
on collection query optimization, such as OptiQL~\citep{Rompf13,Arvind13}, gave
better speedups, as also discussed in \cref{sec:rwdsl-lms}.

@@

Another problem is that most fusion techniques are designed for sequential
processing, hence conflict with incrementalization.
Most fusion techniques generally assume bulk operations scan collections in
linear order. For instance, shortcut fusion rewrites operators in terms of
\texttt{foldr}.
During parallel and/or incremental computation, instead, it is better to use
\emph{tree-shaped folds}: that is, to split the task in a divide-and-conquer
fashion, so that the various subproblems form a balanced tree. This division
minimizes the height of the tree, hence the number of steps needed to combine
results from subproblems into the final result, as also discussed by
\citet{Steele2009organizing}. It is not so clear how to apply shortcut fusion to
parallel and incremental programs.
\citet{Maier2013} describe an incremental tree-shaped fold, where each
subproblem that is small enough is solved by scanning its input in linear order,
but does not perform code transformation and does not study how to perform fusion.
